Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-01 text/back_end_KS.qmd
methods smoothing

**Research design**. We utilize a multiple, quantitative case study design in which metropolitan areas are cases. The research questions of this study aim to explore the socioeconomic and racial variation between Geomarkets, how such variation changes over time, and how Geomarket landscapes shape the composition of included versus excluded prospects. Our goal is not to generalize Geomarket variation nationally, but rather to explore how differences across Geomarkets landscapes shape prospect pools based on the geographical model of metropolitan areas and their demographic characteristics. Such aims align with the purpose of a multiple case study design to gain a deeper understanding of the phenomenon of interest by comparing and contrasting across different settings \[CITE YIN 2001\]. In doing so, multiple cases allow researchers to explore the "how" and "why" of the phenomenon in rich and nuanced contexts. We therefore focus on minimizing the number of metropolitan areas in order to provide sufficient analytic depth while also situating each case within its specific historic and contemporary geographical landscape \[TIE IN THEORETICAL FRAMEWORK HERE A BIT MORE\]. Online Appendix XXX presents results for a larger number of metropolitan areas.

Case selection was informed by based on purposeful sampling to identify “information-rich” cases (Patton, 2002; Yin, 2009) that can provide in-depth knowledge about the phenomenon of interest. We first considered metropolitan areas as information rich cases across two broad dimensions: 1) Geographical and urban model structures and 2) variation across racial/ethnic and socioeconomic characteristics. Based on our read of @RN4982, we are interested in "major metropolitan area\[s\] \ldots composed of several markets, usually corresponding to the inner city, a first ring of suburbs, and an outer ring of suburbs" (p. 11-12). Building on our conceptual framework, we are interested in racial and socioeconomic inequality between geomarkets within a metropolitan area (RQ1) and how this inequality contributes to inequality in which prospective students are targeted by student list purchases (RQ2). \[DISCUSSION OF WHICH CASES TO SELECT. KARINA\]. Choice of metropolitan areas is informed by data availability. Although Census data are available for all metropolitan areas (RQ1) we do not have good candidate student list purchases for all metropolitan areas (RQ2).

Then, our purposeful sampling approach also took into consideration student list orders! \[SAY SOMETHING ABOUT WHAT IS A GOOD CANDIDATE STUDENT LIST PURCHASE -- PURCHASE WITH NOT TOO MANY CRITERIA THAT INCLUDE ALL GEOMARKETS IN A METRO\].

>>>>

**Research design**. We utilize a multiple, quantitative case study design in which metropolitan areas are cases. Case studies are used to gain a deeper understanding of the phenomenon of interest by comparing and contrasting across different settings \[CITE YIN 2001\]. Our study's research questions explore the socioeconomic and racial variation between Geomarkets, how such variation changes over time, and how Geomarket landscapes shape the composition of included versus excluded prospects. Our goal is not to generalize Geomarket variation nationally, but rather to explore how differences across Geomarket landscapes shape prospect pools based on the geographical model of metropolitan areas and their demographic characteristics. Such aims align with the purpose of case studies, which drive the exploration of research questions focused on the "how" and "why" of phenomenon in rich and nuanced contexts. We therefore focus on minimizing the number of metropolitan areas in order to provide sufficient analytic depth while situating each case within its specific historic and contemporary sociodemographic characteristics \[TIE IN THEORETICAL FRAMEWORK HERE A BIT MORE\]. Online Appendix XXX presents results for a larger number of metropolitan areas.

Case selection was therefore informed by purposeful sampling to identify “information-rich” cases (Patton, 2002; Yin, 2009) that can provide in-depth knowledge about variation across Geomarkets and overtime. We considered metropolitan areas as information rich cases across three broad dimensions: 1) urban model structures, 2) geographical variation, 3) variation across racial/ethnic and socioeconomic characteristics, and 4) availability of student list orders and de-identified list data. Based on our read of @RN4982, we identified classic urban model structures that reflected "major metropolitan area\[s\] \ldots composed of several markets, usually corresponding to the inner city, a first ring of suburbs, and an outer ring of suburbs" (p. 11-12), such as Boston, Chicago, Detroit, Baltimore, Denver, Houston, Miami, Philadelphia**,** Phoenix, and Pittsburgh. To compare and contrast with classic urban model structures, we also identified alternative urban models such as bimodal or metroplex models that contain two centralized cores surrounded by the more classing suburban ring (e.g., Dallas-Fort Worth and San Francisco) and multi-nuclei models where urban growth develops around multiple, sometimes uncentralized and independent, metropolitan areas (e.g., Los Angeles and New York City). The third consideration addresses geographical variations in demographic populations by intentionally selecting metropolitan cases across different U.S. regions (West, Midwest, Northeast, South) to ensure resident and prospect pools mirror broader populations. Cumulatively related to all previous selection dimensions, we are also interested in variation of racial and socioeconomic populations between Geomarkets within a metropolitan areas (RQ1) and how such variation shapes which prospective students are targeted by student list purchases (RQ2). We therefore considered whether or not metropolitan areas provided sufficient and varied racial and socioeconomic diversity for our research purposes.

Lastly, choice of metropolitan cases was informed by student list data availability. Although Census data are available for all metropolitan areas nationally (RQ1), we do not have good candidate student list data for all metropolitan areas (RQ2). In order to simulate which prospective students would have been included or excluded from the student list purchase had the purchase filtered on particular Geomarkets, student list data must contain orders that resulted in a large pool of prospects within the metropolitan areas considered above that would include all Geomarkets within metropolitan areas. This lead us to select on orders that filtered for an entire state (e.g., State filter is set to Pennsylvania), an entire metropolitan area without a Geomarket filter (CBSA filter is set to the Philadelphia metropolitan area) , or an entire metropolitan area with a Geomarket filter inclusive of all Geomarkets (CBSA filter is set to the Philadelphia metropolitan area and Geomarket filter is set to all five Geomarkets in the Philadelphia area). In order to isolate the impact of Geomarkets on the racial/ethnic and socioeconomic characteristics of prospect pools, a suitable metropolitan case candidate would also be contingent on student list orders that did not include too many filtering criteria beyond those detailed previously. Because nearly all orders we collected (Author, BLINDED) filter on some academic criteria, we only considered orders that included assessment threshold filters (e.g., PSAT, SAT) across a wide score range to explore whether and to what extent Geomarket contributions to prospect pools are consistent across all score ranges.

Our case selection process resulted in three metropolitan cases: Chicago, Dallas-Fort Worth, and Los Angeles.
